---
title: "Summary - Maths & Statistics"
output:
  pdf_document: default
  html_document:
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Basics

### 0.1 Managing Packages
```{r}
# install.packages("ggplot2")
library(ggplot2)
```

### 0.2 Data Types
```{r results='hold'}
typeof(5)
typeof("text")
typeof(TRUE)
```

### 0.3 Arithmetic with R
```{r results='hold'}
5 + 5 # An addition
5 - 4 # A subtraction
3 * 5 # A multiplication
9 / 3 # A division
2 ^ 5 # Exponentiation
7 %% 3 # Modulo
```

### 0.4 Variable Assignment
```{r}
var1 <- 5
var2 <- 13
var1 + var2
```
<br><br>  

## 1. Vectors & Matrices

### 1.1 Vectors

#### 1.1.1 Creating a vector
```{r results='hold'}
v1 <- c(1, 2, 3, 4, 5)
v1
length(v1)
```
<br>  

#### 1.1.2 Getting the components of a vector (slicing)
```{r results='hold'}
v1[1]
v1[1:3]
```
<br>  

#### 1.1.3 Auto population of a vector
```{r results='hold'}
c(rep(0, times = 5))
c(rep(c(1, 2), times = 3))
c(5:9)
c(seq(from = 1, to = 7, by = 2))
c(seq(from = 1, to = 7, length = 5))
```
<br>  

#### 1.1.4 Arithmetic with vectors
```{r results='hold'}
a <- 1:5
b <- 3:7
a+b
a-b
a*b
a/b
a^(b-a)
```
And the scalar
```{r}
a%*%b
```
<br><br>  

### 1.2 Matrices

#### 1.2.1 Creating a matrix
```{r}
m1 <- matrix(c(1:6), ncol = 3, byrow = TRUE)
m1
```
```{r}
diag(c(1:4))
diag(3)
```
<br>  

#### 1.2.2 Length, Dimensions and Slicing
```{r results='hold'}
length(m1)
dim(m1)
```
```{r results='hold'}
m1[2,3]
m1[,2]
```
<br>  

#### 1.2.3 Arithmetic with matrices
```{r results='hold'}
m2 <- matrix(c(1:6), ncol = 2, byrow = TRUE)
m1
m2
```
```{r error=TRUE}
3*m1 # scalar multiplication
m1%*%m2
m2%*%m1
m1*m2
```
<br>  

#### 1.2.4 Transpose matrices
Let ${\bf M1}=[a_{ij}]$ then ${\bf M1'}=[a_{ji}]$ 
```{r}
m1
t(m1)
```
$({\bf M1'})'={\bf M1}$
```{r}
t(t(m1))
```
${\bf M1}{\bf M1'}$ always yields a symmetric matrix ${\bf Q}$
```{r}
m1%*%t(m1)
```
<br>  

#### 1.2.5 Inverse of matrix
An $n \times n$ matrix ${\bf M}$ has an inverse, denoted ${\bf M^{-1}}$, provided that ${\bf M}{\bf M^{-1}}={\bf I_{n}}$
```{r}
m3 <- matrix(c(2, 5, 1, 3), nrow = 2, byrow = TRUE)
m3
solve(m3)
m3%*%solve(m3)
```
Otherwise it is said to be noninvertible or singular:
```{r error=TRUE}
solve(matrix(c(1:9), nrow = 3))
```
<br>  

#### 1.2.6 Rank of a matrix
```{r}
m4 <- matrix(c(1,2,3,2,4,6,10,20,30), ncol = 3, byrow = FALSE)
m4
qr(m4)$rank
```
The rank of the matrix is 1, because $Col2 = 2*Col1$ and $Col3 = 10*Col1$. This attribute can be used to reduce the size of matrices and data.
<br>  

#### 1.2.7 Eigenvalues
We use the example of a $3 \times 3$ matrix. This could be considered as the variance-covariance matrix of three variables, but the main thing is that the matrix is <strong>square</strong> and <strong>symmetric</strong>, which guarrantees that the eigenvalues $\lambda_{i}$ are real numbers. 
```{r}
m5 <- matrix(c(13, -4, 2, -4, 11, -2, 2, -2, 8), ncol = 3, byrow = TRUE)
m5
```
```{r}
eigen(m5)
```
As shown above, this returns a namend list, containing the eigenvalues and eigenvectors.
One may test the result by calculating the required equation for this eigenvalues:<br>
<center>$(A- \lambda I_{n})x=0$</center><br>
which leads to:
<center>$det(A- \lambda I_{n})=0$</center><br>  
```{r}
m_test1 = m5 - eigen(m5)$values[1] * diag(3)
m_test1
round(det(m_test1), digits = 2)
```
Which should also be true for the second eigenvalue:
```{r}
m_test2 = m5 - eigen(m5)$values[2] * diag(3)
round(det(m_test2), digits = 2)
```